A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer

### 2024-01-16 content/manuscript.qmd
correct/expand already written stuff

---
title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliations:
  - id: 1
    name: University of Zurich
    departement: Departement of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    departement: Departement of Radiation Oncology
    city: Zurich
    country: Switzerland
bibliography: ../references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
format:
  iccr-pdf:
    keep-tex: true
    include-in-header:
      text: "\\usepackage{newtxtext,newtxmath}"
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999]. Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021]. Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease. Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@vorwerk_guidelines_2011;@gregoire_delineation_2018;@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999;@chao_determination_2002].

@@

2. an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease [@ludwig_hidden_2021;@ludwig_modelling_2023], given an individual patient's diagnosis.

@@

HNSCC patients with primary tumors at different subsites, e.g. in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@candela_patterns_1990;@woolgar_histological_1999]. To naively capture this in a risk prediction using our previously developed model, one would need to train entirely separate models for each subsite, based on separated datasets. We consider this a limitation for two reasons:

@@

Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a hidden binary random variable (RV) $X_v[t]$ representing the true -- but not directly observable -- state of that level at the abstract time-step $t$. It may take on the values "healthy" (0) or "involved" (1). Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's hidden state w.r.t. their lymphatic involvement at time $t$.

@@

![Example of a parametrized graph representation of a small lymphatic network consisting of only three LNLs. Blue, round nodes represent the hidden RVs, while the orange square nodes show the observed RVs. Every arc represents a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.png){#fig-graph-with-obs width="25%"}

@@

: Conditional probability $P\left( X_3[t+1] \mid X_2[t] \right)$ for a spread from LNL II to III during the transition from $t$ to $t+1$. This corresponds to one of the blue arcs in @fig-graph-with-obs. {#tbl-spread-2-to-3 tbl-colwidths="[25,18,40]"}

@@

And set every entry of this starting distribution to zero, except the first entry, which we set to one. This means at $t=0$ there is a probability of one to be in the completly healthy state $\boldsymbol{\xi}_0 = \begin{pmatrix} 0, 0, \ldots, 0 \end{pmatrix}$.

@@

$$
P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid T \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
$$ {#eq-marginalized-evolution}

@@

Note that [the above equation](#eq-marginalized-evolution) still depends on the parametrization of the transition matrix. We collect these parameters in a vector $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v}, p_\text{advanced} \}$.

To train our model, we need to compute the likelihood of a dataset $\mathcal{D}=\left( \mathbf{Y}_1, \mathbf{Y}_2, \ldots, \mathbf{Y}_N \right)$ of $N$ previously observed patient diagnoses, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the sums in the denominator of @eq-bayes-law:

$$
\mathcal{L}(\mathcal{D} \mid \boldsymbol{\theta}) = \prod_{i=1}^N \sum_{j=1}^{2^V} P \left( \mathbf{Y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i \right)
$$

@@

The just introduced model is capable of learning one set of or distribution over parameters $\boldsymbol{\theta}_s$ from a cohort of patients with a primary tumor in a given subsite $s$. If we tried to train it with a cohort consisting of patients with tumors in two very different subsites, the model would likely learn parameters that represent a compromise between the two subcohort's true parameters. This compromise might describe neither of the subcohorts' lymphatic spread patterns sufficiently well.

@@

$$
P \left( \mathbf{X} \mid \boldsymbol{\Psi} \right) = \sum_{i=0}^g \pi_i \cdot P \left( \mathbf{X} \mid \boldsymbol{\theta}_i \right)
$$ {#eq-mixture-distribution}

@@

1. We draw a categorical RV $C_j$ that may take on the values $1, \ldots, g$ with corresponding probabilities $\pi_1, \ldots, \pi_g$:
  $$
  P(C_j=c_j) = \pi_{c_j}
  $$
  For convenience, we also transform this into a one-hot encoding $\mathbf{z}_j$ that is zero everywhere, except $z_{ij}=1$ when $C_j=i$:
  $$
  P \left( \mathbf{Z}_j=\mathbf{z}_j \right) = \prod_{i=0}^g \pi_i^{z_{ij}}
  $$

2. Then, we draw $\mathbf{y}_j$ from the conditional probability of $P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right)$, which is now given by a slightly modified version of @eq-mixture-distribution:
  $$
  P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right) = \frac{\sum_{i=0}^g z_{ij} \cdot \pi_i \cdot P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_i \right)}{\sum_{i=0}^g z_{ij} \cdot \pi_i} = P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_\ell \right)
  $$
  where $\ell$ is just so that $z_{\ell j}=1$, and all other entries of $\mathbf{z}_j$ are zero.

>>>>

---
title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliations:
  - id: 1
    name: University of Zurich
    departement: Departement of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    departement: Departement of Radiation Oncology
    city: Zurich
    country: Switzerland
bibliography: ../references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
format:
  iccr-pdf:
    keep-tex: true
    include-in-header:
      - text: "\\usepackage{newtxtext,newtxmath}"
  html:
    code-line-numbers: true
    code-tools: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@woolgar_histological_1999]. Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021]. Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease. Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@vorwerk_guidelines_2011;@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].

@@

2. an interpretable hidden Markov model (HMM), trained with this data, to predict the risk for occult nodal disease [@ludwig_hidden_2021], given an individual patient's diagnosis.

@@

HNSCC patients with primary tumors at different subsites, e.g. in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999]. To naively capture this in a risk prediction using our previously developed model, one would need to train entirely separate models for each subsite, based on separated datasets. We consider this a limitation for two reasons:

@@

Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a hidden binary random variable (RV) $X_v[t]$ taking on the true -- but not directly observable -- state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved"). Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's hidden state w.r.t. their lymphatic involvement at time $t$.

@@

![Parametrized graph representation of the lymphatic network considering four LNLs. Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs. Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="25%"}

@@

: Conditional probability $P\left( X_3[t+1] \mid X_2[t] \right)$ for a spread from LNL II to III during the transition from $t$ to $t+1$. This corresponds to one of the blue arcs in @fig-graph-with-obs. Note that the values in the row with $X_3[t+1]=0$ is all zeros, and the row with $X_3[t+1]=1$ all ones if $X_3[t]=1$. {#tbl-spread-2-to-3 tbl-colwidths="[25,18,40]"}

@@

And set every entry of this starting distribution to zero, except the first one, which we set to one. This means at $t=0$ there is a probability of one to be in the completly healthy state $\boldsymbol{\xi}_0 = \begin{pmatrix} 0, 0, \ldots, 0 \end{pmatrix}$.

@@

$$
P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid T, \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
$$ {#eq-marginalized-evolution}

@@

Note that @eq-marginalized-evolution still depends on the parametrization of the transition matrix. We collect these parameters in a vector $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v}, p_\text{advanced} \}$.

To train our model, we need to compute the likelihood of a dataset $\mathcal{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ previously observed patient diagnoses, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the sums in the denominator of @eq-bayes-law:

$$
\mathcal{L}(\mathcal{D} \mid \boldsymbol{\theta}) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
$$

@@

The just introduced model is capable of learning one set of or distribution over parameters $\boldsymbol{\theta}$ from a cohort of patients with a primary tumor in a given subsite. If we tried to train it with a cohort consisting of patients with tumors in two very different subsites, the model would likely learn parameters that represent a compromise between the two subcohort's true parameters. This compromise might describe neither of the subcohorts' lymphatic spread patterns sufficiently well.

@@

$$
P \left( \mathbf{Y} \mid \boldsymbol{\Psi} \right) = \sum_{i=0}^g \pi_i \cdot P \left( \mathbf{Y} \mid \boldsymbol{\theta}_i \right)
$$ {#eq-mixture-distribution}

@@

1. We draw a categorical RV $C_j$ that may take on the values $1, \ldots, g$ with corresponding probabilities $\pi_1, \ldots, \pi_g$. Hence, we have $P(C_j=c_j) = \pi_{c_j}$. For convenience, we also transform this into a one-hot encoding $\mathbf{z}_j$ that is zero everywhere, except $z_{\ell j}=1$ when $C_j=\ell$:
  $$
  P \left( \mathbf{Z}_j=\mathbf{z}_j \right) = \prod_{i=0}^g \pi_i^{z_{ij}}
  $$

2. Then, we draw $\mathbf{y}_j$ from the conditional probability of $P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right)$, which is now given by a slightly modified version of @eq-mixture-distribution:
  $$
  P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right) = \frac{\sum_{i=0}^g z_{ij} \pi_i P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_i \right)}{\sum_{i=0}^g z_{ij} \pi_i} = P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_\ell \right)
  $$
  where $\ell$ is just so that $z_{\ell j}=1$, and all other entries of $\mathbf{z}_j$ are zero.
